### 2025-03-10 paper/preprint/paper.tex
done with grid search appendix

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g.~cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: 1) the contrastive
divergence, \(\text{div}(\cdot)\), between counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the target class. We generate samples
\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
that target over \(T\) iterations using a gradient-based counterfactual
generator. This is similar to how conditional sampling is used to draw
from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
(\citeproc{ref-grathwohl2020your}{2020}).

@@

@@

@@

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note}, rightrule=.15mm, titlerule=0mm]

@@

We begin by summarizing the high-level findings in
Section~\ref{sec-app-grid-hl}. For each of the categories,
Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
present all details including the parameter grids, average performance
outcomes and key evaluation metrics for the generated counterfactuals.

\subsubsection{High-Level Findings}\label{sec-app-grid-hl}

Overall, we find that Counterfactual Training (CT) achieves it key
objectives consistently across all hyperparameter settings and also
broadly across datasets. We do observe strong sensitivity to certain
hyperparameters, but clear and manageable patterns emerge in those
cases. With respect to the underlying training data distribution, we
find that outcomes for CT seem to depend to some degree on class
separability.

@@

\paragraph{Key Counterfactual
Outcomes}\label{key-counterfactual-outcomes}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g.~cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: 1) the contrastive
divergence, \(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(\mathbf{x}\) and, 2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
particular, we set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
a mature counterfactual for a randomly sampled factual from a non-target
class generated through a gradient-based counterfactual generator for at
most \(T\) iterations. We define mature counterfactuals as those that
have either exhausted \(T\) or reached convergence in terms of the
pre-determined decision threshold earlier.

@@

\subsubsection{Evaluation}\label{evaluation}

Our key outcome of interest is how well models perform with respect to
explainability (Definition~\ref{def-explainability}): to this end, we
focus primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain. We rely on the distance-based metric used by Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}),

\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation}

and introduce a novel divergence metric,

\begin{equation}\phantomsection\label{eq-impl-div}{
\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation}

where \(\mathbf{X}^\prime\) denotes a set of multiple counterfactuals
and \(\text{MMD}(\cdot)\) is an unbiased estimate of the squared
population maximum mean discrepancy
(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
Equation~\ref{eq-impl-div} is equal to zero iff
\(\mathbf{X}^\prime=\mathbf{X}^+\).

In addition to cost and plausibility, we also compute other standard
metrics to evaluate counterfactuals at test time including validity and
redundancy. Finally, we also assess the predictive performance of models
using standard metrics.

@@

\bibitem[\citeproctext]{ref-gretton2012kernel}
Gretton, Arthur, Karsten M Borgwardt, Malte J Rasch, Bernhard Schölkopf,
and Alexander Smola. 2012. {``A Kernel Two-Sample Test.''} \emph{The
Journal of Machine Learning Research} 13 (1): 723--73.

@@

\subsection{Other Technical Details}\label{other-technical-details}

\begin{equation}\phantomsection\label{eq-mmd}{
\begin{aligned}
MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
\end{aligned}
}\end{equation}

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

@@

We begin by summarizing the high-level findings in
Section~\ref{sec-app-grid-hl}. For each of the categories,
Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
present all details including the exact parameter grids, average
predictive performance outcomes and key evaluation metrics for the
generated counterfactuals.

\subsubsection{Evaluation Details}\label{evaluation-details}

To measure predictive performance, we compute the accuracy and F1-score
for all models on test data (Table~\ref{tbl-acc-gen},
Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
explanatory performance, we report here our findings for the
(im)plausibility and cost of counterfactuals at test time. Since the
computation of our proposed divergence metric
(Equation~\ref{eq-impl-div}) is memory-intensive, we rely on the
distance-based metric for the grid searches. For the counterfactual
evaluation, we draw factual samples from the training data for the grid
searches to avoid data leakage with respect to our final results
reported in the body of the paper. Specifically, we want to avoid
choosing our default hyperparameters based on results on the test data.
Since we are optimizing for explainability, not predictive performance,
we still present test accuracy and F1-scores.

@@

\paragraph{Counterfactual Outcomes}\label{sec-app-grid-hl}

Overall, we find that Counterfactual Training (CT) achieves it key
objectives consistently across all hyperparameter settings and also
broadly across datasets: plausibility is improved by up to
\textasciitilde60 percentage points (ppts) for the \emph{Circles} data
(e.g. Figure~\ref{fig-grid-gen_params-plaus-circles}),
\textasciitilde25-30ppts for the \emph{Moons} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-moons}) and
\textasciitilde10-20ppts for the \emph{Linearly Separable} data (e.g.
Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
average costs of faithful counterfactuals are reduced in many cases by
around \textasciitilde20-25ppts for \emph{Circles} (e.g.
Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
\textasciitilde50ppts for \emph{Moons} (e.g.
Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
Separable} data, costs are generally increased although typically by
less than 10ppts (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
which reflects a common tradeoff between costs and plausibility
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).

We do observe strong sensitivity to certain hyperparameters, with clear
an manageable patterns. Concerning generator parameters, we firstly find
that using \emph{REVISE} to generate counterfactuals during training
typically yields the worst outcomes out of all generators, often leading
to a substantial decrease in plausibility. This finding can be
attributed to the fact that \emph{REVISE} effectively assigns the task
of learning plausible explanations from the model itself to a surrogate
VAE. In other words, counterfactuals generated by \emph{REVISE} are less
faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
would expect them to be a less effective and, in fact, potentially
detrimental role in our training regime. Secondly, we observe that
allowing for a higher number of maximum steps \(T\) for the
counterfactual search generally yields better outcomes. This is
intuitive, because it allows more counterfactuals to reach maturity in
any given iteration. Looking in particular at the results for
\emph{Linearly Separable}, it seems that higher values for \(T\) in
combination with higher decision thresholds (\(\tau\)) yields the best
results when using \emph{ECCo}. But depending on the degree of class
separability of the underlying data, a high decision-threshold can also
affect results adversely, as evident from the results for the
\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
here we find that CT generally fails to achieve its objective because
only a tiny proportion of counterfactuals ever reaches maturity.

Regarding penalty strengths, we find that the strength of the energy
regularization, \(\lambda_{\text{reg}}\) is a key hyperparameter, while
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. In particular, we observe
that not regularizing energy enough or at all typically leads to poor
performance in terms of decreased plausibility and increased costs, in
particular for \emph{Circles} (Figure~\ref{fig-grid-pen-plaus-circles}),
\emph{Linearly Separable} (Figure~\ref{fig-grid-pen-plaus-lin_sep}) and
\emph{Overlapping} (Figure~\ref{fig-grid-pen-plaus-over}). High values
of \(\lambda_{\text{reg}}\) can increase the variability in outcomes, in
particular when combined with high values for \(\lambda_{\text{div}}\)
and \(\lambda_{\text{adv}}\), but this effect is less pronounced.

Finally, concerning other hyperparameters we observe that the
effectiveness and stability of CT is positively associated with the
number of counterfactuals generated during each training epoch, in
particular for \emph{Circles}
(Figure~\ref{fig-grid-train-plaus-circles}) and \emph{Moons}
(Figure~\ref{fig-grid-train-plaus-moons}). We further find that a higher
number of training epochs is beneficial as expected, where we tested
training models for 50 and 100 epochs. Interestingly, we find that it is
not necessary to employ CT during the entire training phase to achieve
the desired improvements in explainability: specifically, we have tested
training models conventionally during the first half of training before
switching to CT after this initial burn-in period.

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]

### 2025-03-10 paper/sections/method.qmd
done with grid search appendix

where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g. cross-entropy). The two additional components in @eq-obj are explained in more detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

@@

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the target class. We generate samples $\mathbf{x}_T^\prime$ by first randomly sampling the target class $y^+ \sim p(y)$ and then generating a counterfactual explanation for that target over $T$ iterations using a gradient-based counterfactual generator. This is similar to how conditional sampling is used to draw from $p_\theta(\mathbf{x})$ in @grathwohl2020your.

>>>>

where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g. cross-entropy). The two additional components in @eq-obj are explained in more detail below. For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

@@

where $\mathcal{E}_\theta(\cdot)$ denotes the energy function. In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed data point sampled from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$. Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated through a gradient-based counterfactual generator for at most $T$ iterations. We define mature counterfactuals as those that have either exhausted $T$ or reached convergence in terms of the pre-determined decision threshold earlier.

### 2025-03-10 paper/sections/other/grids.qmd
done with grid search appendix

To measure predictive performance, we compute the accuracy and F1-score for all models on test data (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train).

We evaluated

### High-Level Findings {#sec-app-grid-hl}

Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved  We do observe strong sensitivity to certain hyperparameters, but clear and manageable patterns emerge in those cases. With respect to the underlying training data distribution, we find that outcomes for CT seem to depend to some degree on class separability.

@@

#### Key Counterfactual Outcomes

>>>>

To measure predictive performance, we compute the accuracy and F1-score for all models on test data (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train). With respect to explanatory performance, we report here our findings for the (im)plausibility and cost of counterfactuals at test time. Since the computation of our proposed divergence metric (@eq-impl-div) is memory-intensive, we rely on the distance-based metric for the grid searches. For the counterfactual evaluation, we draw factual samples from the training data for the grid searches to avoid data leakage with respect to our final results reported in the body of the paper. Specifically, we want to avoid choosing our default hyperparameters based on results on the test data. Since we are optimizing for explainability, not predictive performance, we still present test accuracy and F1-scores.

@@

#### Counterfactual Outcomes {#sec-app-grid-hl}

Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percentage points (ppts) for the *Circles* data (e.g. @fig-grid-gen_params-plaus-circles), ~25-30ppts for the *Moons* data (e.g. @fig-grid-gen_params-plaus-moons) and ~10-20ppts for the *Linearly Separable* data (e.g. @fig-grid-gen_params-plaus-lin_sep). At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25ppts for  *Circles* (e.g. @fig-grid-gen_params-cost-circles) and up to ~50ppts for *Moons* (e.g. @fig-grid-gen_params-cost-moons). For the *Linearly Separable* data, costs are generally increased although typically by less than 10ppts (e.g. @fig-grid-gen_params-cost-lin_sep), which reflects a common tradeoff between costs and plausibility [@altmeyer2024faithful].

We do observe strong sensitivity to certain hyperparameters, with clear an manageable patterns. Concerning generator parameters, we firstly find that using *REVISE* to generate counterfactuals during training typically yields the worst outcomes out of all generators, often leading to a substantial decrease in plausibility. This finding can be attributed to the fact that *REVISE* effectively assigns the task of learning plausible explanations from the model itself to a surrogate VAE. In other words, counterfactuals generated by *REVISE* are less faithful to the model that *ECCo* and *Generic*, and hence we would expect them to be a less effective and, in fact, potentially detrimental role in our training regime. Secondly, we observe that allowing for a higher number of maximum steps $T$ for the counterfactual search generally yields better outcomes. This is intuitive, because it allows more counterfactuals to reach maturity in any given iteration. Looking in particular at the results for *Linearly Separable*, it seems that higher values for $T$ in combination with higher decision thresholds ($\tau$) yields the best results when using *ECCo*. But depending on the degree of class separability of the underlying data, a high decision-threshold can also affect results adversely, as evident from the results for the *Overlapping* data (@fig-grid-gen_params-plaus-over): here we find that CT generally fails to achieve its objective because only a tiny proportion of counterfactuals ever reaches maturity.

Regarding penalty strengths, we find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is a key hyperparameter, while sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. In particular, we observe that not regularizing energy enough or at all typically leads to poor performance in terms of decreased plausibility and increased costs, in particular for *Circles* (@fig-grid-pen-plaus-circles), *Linearly Separable* (@fig-grid-pen-plaus-lin_sep) and *Overlapping* (@fig-grid-pen-plaus-over). High values of $\lambda_{\text{reg}}$ can increase the variability in outcomes, in particular when combined with high values for $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$, but this effect is less pronounced.

Finally, concerning other hyperparameters we observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch, in particular for *Circles* (@fig-grid-train-plaus-circles) and *Moons* (@fig-grid-train-plaus-moons). We further find that a higher number of training epochs is beneficial as expected, where we tested training models for 50 and 100 epochs. Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability: specifically, we have tested training models conventionally during the first half of training before switching to CT after this initial burn-in period.
